---
title: "Gun Deaths Analysis"
author: "Owen Morales"
format:
  html:
    code-fold: true
    code-line-numbers: true
execute:
  error: true
---

## Setup

I will create a new `season` variable based on the `month` column to address the client's request for seasonal commercials

```{r setup}

library(tidyverse)

guns <- read_csv("https://raw.githubusercontent.com/fivethirtyeight/guns-data/master/full_data.csv")

gunsClean <- guns %>%
  filter(!is.na(intent)) %>%
  mutate(
    month_num = as.numeric(month),
    season = case_when(
      month_num %in% c(12, 1, 2) ~ "Winter",
      month_num %in% c(3, 4, 5) ~ "Spring",
      month_num %in% c(6, 7, 8) ~ "Summer",
      month_num %in% c(9, 10, 11) ~ "Fall"
    ),
    season = factor(season, levels = c("Winter", "Spring", "Summer", "Fall"))
  )

```

The client needs to know which demographic groups to target based on the overall causes of gun deaths.

```{r}

gunsClean %>%
  count(intent, race) %>%
  ggplot(aes(x = reorder(intent, n), y = n, fill = race)) +
  geom_col() +
  coord_flip() +
  theme_minimal() +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Total Gun Deaths by Intent and Race (2012-2014)", x = "Intent", y = "Total Deaths", fill = "Race")

```

This plot illustrates that the vast majority of gun deaths are driven by two primary intents: suicides and homicides. It highlights a stark demographic divide that the client must consider for targeting: suicides are overwhelmingly committed by White individuals, whereas homicides disproportionately affect Black individuals (though we should consider that the white population is bigger than any of the other demographics we have seen). Targeting these specific demographic groups based on the intent of the messaging is crucial for any overall intervention strategy

# Seasonal Commercial Targeting

The client wants to create commercials tailored to different seasons

```{r}

gunsClean %>%
  count(season, intent) %>%
  ggplot(aes(x = season, y = n, group = intent, color = intent)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  theme_minimal() +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Gun Deaths Across Seasons by Intent", x = "Season", y = "Total Deaths", color = "Intent")

```

This visualization tracks the primary intents of gun deaths across the four seasons to guide the timing of different ad campaigns. It reveals that homicides experience a noticeable spike during the summer months, while suicides peak slightly in the spring and summer before declining in the winter. This suggests that anti-homicide campaigns should be heavily funded and broadcast just before and during the summer season

```{r}

gunsClean %>%
  filter(intent == "Homicide", season == "Summer") %>%
  ggplot(aes(x = age, fill = sex)) +
  geom_histogram(binwidth = 3, position = "identity", alpha = 0.7) +
  theme_minimal() +
  labs(
    title = "Summer Homicides by Age and Sex", x = "Age", y = "Count of Victims", fill = "Sex")

```

Focusing specifically on the summer surge in homicides, this histogram shows the distribution of victims by age and sex. Young males, primarily between the ages of 18 and 30, constitute the absolute largest demographic group of victims during this high-risk season. Summer commercials aimed at reducing community violence should feature messaging, actors, and channels heavily tailored toward young men

```{r}

gunsClean %>%
  filter(intent == "Suicide") %>%
  ggplot(aes(x = age, fill = sex)) +
  geom_density(alpha = 0.6) +
  facet_wrap(~season) +
  theme_minimal() +
  labs(title = "Age Distribution of Suicides by Season and Sex", x = "Age", y = "Density", fill = "Sex")

```

This density plot breaks down the age distribution of suicides by season. It reveals that while suicides occur across all age groups, there is a noticeable concentration among middle-aged individuals (ages 40-60) in the spring and summer months.

