---
title: "Combining Height Datasets"
author: "Owen Morales"
format:
  html:
    code-fold: true
    code-line-numbers: true
execute:
  error: true
---

## 1

```{r message=FALSE, warning=FALSE}

library(haven)   # reading .dta (stata) and .sav (SPSS) files
library(foreign) # reading .dbf files
library(readr)   # reading .csv files
library(dplyr)   
library(ggplot2)

# 1. German male conscripts in Bavaria (.dta)
germanConscr <- read_dta("https://byuistats.github.io/M335/data/heights/germanconscr.dta")

# 2. Heights of Bavarian male conscripts (.dta)
germanPrison <- read_dta("https://byuistats.github.io/M335/data/heights/germanprison.dta")

# 3. Heights of south-east and south-west German soldiers (.zip containing .dbf)
# Create a temporary file and a temporary directory to handle the zip
tempZip <- tempfile(fileext = ".zip")
tempDir <- tempdir()

# Download the zip file (using mode = "wb" for binary files)
download.file("https://byuistats.github.io/M335/data/heights/Heights_south-east.zip", 
              destfile = tempZip, mode = "wb")

# Unzip into the temporary directory
unzip(tempZip, exdir = tempDir)

# Find the .dbf file that was just unzipped and read it
dbfFile <- list.files(tempDir, pattern = "\\.DBF$|\\.dbf$", full.names = TRUE)
germanSoldiers <- read.dbf(dbfFile[1])

# 4. Bureau of Labor Statistics Height Data (.csv)
# We will load it, filter for males (the column is named 'sex' in this dataset), 
# and add the birth year as 1950.
blsDataRaw <- read_csv("https://raw.githubusercontent.com/hadley/r4ds/main/data/heights.csv")

blsData <- blsDataRaw %>%
  filter(sex == "male") %>%
  mutate(birthYear = 1950)

# 5. University of Wisconsin National Survey Data
wisconsinSurvey <- read_sav("https://www.ssc.wisc.edu/nsfh/wave3/NSFH3%20Apr%202005%20release/main05022005.sav")

```

## 2

```{r}

# 1. Bavarian Conscripts
germanConscr_clean <- germanConscr %>%
  mutate(
    birth_year = bdec, 
    height.cm = height,          
    height.in = height.cm / 2.54, # cm -> inch
    study = "Bavarian Conscripts") %>%
  select(birth_year, height.in, height.cm, study)

# 2. Bavarian Prisoners
germanPrison_clean <- germanPrison %>%
  mutate(
    birth_year = bdec,
    height.cm = height,         
    height.in = height.cm / 2.54, # cm -> inch
    study = "Bavarian Prisoners") %>%
  select(birth_year, height.in, height.cm, study)

# 3. German Soldiers (18th Century)
germanSoldiers_clean <- germanSoldiers %>%
  mutate(
    birth_year = GEBJ,           # GEBJ = Geburtsjahr (Birth Year)
    height.cm = CMETER,          
    height.in = height.cm / 2.54, # cm -> inch
    study = "German Soldiers") %>%
  select(birth_year, height.in, height.cm, study)

# 4. Wisconsin National Survey
wisconsinSurvey_clean <- wisconsinSurvey %>%
  # Filter out the -1 and -2 "Refused/Don't Know" responses so they don't skew the math
  filter(RT216F > 0 & RT216I >= 0) %>%
  mutate(
    birth_year = DOBY,           # doby = Date of Birth, Year
    height.in = (RT216F * 12) + RT216I, # Combine feet and inches into total inches
    height.cm = height.in * 2.54,       # cm -> inch
    study = "Wisconsin Survey") %>%
  select(birth_year, height.in, height.cm, study)

# 5. Bureau of Labor Statistics
blsData_clean <- blsData %>%
  # We already added birth_year as 1950 in the previous step, but just making sure
  mutate(
    birth_year = 1950,
    height.in = height, 
    height.cm = height.in * 2.54,
    study = "BLS Height Data") %>%
  select(birth_year, height.in, height.cm, study)

# Check the result
head(blsData_clean)

# all 5 datasets into one master dataset
allHeights <- bind_rows(blsData_clean, germanConscr_clean, germanPrison_clean, germanSoldiers_clean, wisconsinSurvey_clean)

```

## 4 (How did I tidy it up?)

To get all these datasets playing nicely together I had to wrangle five totally different file formats (.csv, .dta, .sav, and .dbf) into one clean master dataset. I actually leaned on Gemini to help speed up writing the dplyr logic, which was a huge timesaver! The biggest hurdle was standardizing the units, like the historical German datasets used centimeters, while the modern US ones used inches (or a messy mix of feet and inches). To fix that I just calculated standard height.in and height.cm columns across the board. I also had to make a few hard calls on what data to drop. For the BLS set I filtered out the female participants to keep things apples-to-apples with the male-only German soldier and prisoner records. Finanlly, I scrubbed the Wisconsin survey to drop anyone who answered '-1' or '-2' ('Refused' or 'Don't Know'), because if I hadn't caught that, those negative numbers would have completely wrecked the height conversions and skewed the rest of the analysis. After all that, I used bind_rows to stitch everything together into one big tidy dataset.

## 5

```{r}

# chronological order of the studies using factors (c())
allHeights <- allHeights %>%
  mutate(study = factor(study, levels = c(
    "German Soldiers",       # 18th Century
    "Bavarian Prisoners",    # 19th Century
    "Bavarian Conscripts",   # 19th Century
    "BLS Height Data",       # Mid-20th Century
    "Wisconsin Survey"       # Late 20th/21st Century
  )))

# realistic human heights 
allHeights_filtered <- allHeights %>%
  filter(height.in >= 48 & height.in <= 90)

ggplot(allHeights_filtered, aes(x = study, y = height.in, fill = study)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  theme_minimal() +
  labs(title = "Distribution of Heights Across Centuries",subtitle = "Comparing 18th/19th century German records to modern US surveys", x = "Study (Chronological Order)", y = "Height (Inches)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none" )

```

## 6

When you look at the plot I just generated, the story it tells seems pretty straightforward at first, but then it throws a massive curveball that seems to contradict everything. If you just look at the progression from the 18th and 19th-century German records to the mid-20th-century BLS data, the median height clearly trends upward, jumping from around 65-67 inches to about 70 inches. But then we hit the Wisconsin Survey data from the late 20th century, and the median suddenly tanks back down to around 66 inches, contradicting the whole "up-and-to-the-right" growth trend. If you were just glancing at the chart, you’d think humans peaked in the 1950s and then started shrinking. But the contradiction makes perfect sense when you look under the hood at the data dictionary. The German military/prison records and the BLS dataset we filtered were strictly male-only samples. The Wisconsin survey, however, didn't have a gender identifier, meaning it's a mix of both men and women. Because women have a lower average height, throwing them into the dataset naturally drags the median down and widens the overall spread of the boxplot.

So, if someone asserted that "humans are getting taller over time" based strictly on these datasets, I'd tell them they are technically right, but their data hygiene is a little sloppy. You can clearly see the historical growth trend when comparing the historic male-only datasets to the modern male-only BLS dataset as nutrition, medicine, and living conditions improved over the centuries, average heights went up. But you simply can't use the Wisconsin survey to prove or disprove that specific trend because it's an apples-to-oranges comparison. Overall, my stance is that while the data absolutely supports the idea that human height increased over the last few hundred years, this whole exercise is a massive warning about why you can't just blindly bind_rows() on random datasets and trust the visualization without understanding exactly who is in your sample.

(Shoutout to Gemini for helping me debug the R code, figure out the missing variables, and wrangle these messy datasets into shape to build out this analysis!)