---
title: "Open Data Source Evaluation"
author: "Owen Morales"
format:
  html:
    code-fold: true
    code-line-numbers: true
execute:
  error: true
---

## Potential Data Sources

Here are 4 potential free data sources I explored for this project, along with an assessment of their general quality:

1.  **[Data.gov](https://data.gov/)**: The official US government open data portal. 
    * **Quality**: Highly authoritative but heavily variable in structure. Datasets often require extensive cleaning and formatting, and metadata can sometimes be sparse.
2.  **[Kaggle Datasets](https://www.kaggle.com/datasets)**: A community-driven platform for data science.
    * **Quality**: Highly variable. Some datasets are pristine and meticulously documented by organizations, while others are uploaded by users with no data dictionary, missing values, and unverified sources. 
3.  **[World Bank Open Data](https://data.worldbank.org/)**: Global economic, demographic, and health data.
    * **Quality**: Excellent. The data is rigorously standardized and heavily documented, though you will often encounter missing values for certain developing nations or specific years.
4.  **[TidyTuesday Project](https://github.com/rfordatascience/tidytuesday)**: A weekly social data project in R.
    * **Quality**: Very good. The data is pre-curated to be interesting and accessible, but it is deliberately left slightly "messy" so users can practice their data wrangling skills.
    
## Palmer Penguins

This project focuses on the **Palmer Penguins dataset** (read directly from a raw GitHub URL). 

**Original Question Idea:** *How do the physical characteristics of penguins differ across species and locations in the Palmer Archipelago?*



```{r message=FALSE, warning=FALSE}

library(tidyverse)

url <- "https://raw.githubusercontent.com/allisonhorst/palmerpenguins/main/inst/extdata/penguins.csv"
penguins_raw <- read_csv(url)

penguins_clean <- penguins_raw %>%
  filter(!is.na(species), !is.na(island)) %>%
  mutate(across(where(is.character), as.factor))

glimpse(penguins_clean)

```

## Quality Check #1

Counting NAs for numeric columns

```{r message=FALSE, warning=FALSE}

penguins_clean %>%
  select(where(is.numeric)) %>%
  gather(key = "variable", value = "value") %>%
  mutate(is_missing = is.na(value)) %>%
  group_by(variable) %>%
  summarize(missing_count = sum(is_missing)) %>%
  ggplot(aes(x = reorder(variable, missing_count), y = missing_count)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Data Quality: Missing Values per Variable", x = "Variable", y = "Number of Missing Values") +
  theme_minimal()

```

Observation: We have a small handful of missing values for physical measurements (like flipper length and bill depth) but it is a very small percentage of the total dataset. Sex is missing more frequently, which we will need to keep in mind

## Quality Check #2

Distribution of body mass to ensure no extremes

```{r message=FALSE, warning=FALSE}

ggplot(penguins_clean, aes(x = body_mass_g)) +
  geom_histogram(fill = "coral", color = "black", bins = 30) +
  labs(title = "Data Quality: Distribution of Body Mass", x = "Body Mass (grams)", y = "Count") +
  theme_minimal()

```

Observation: The distribution of body mass appears to be reasonably normal with no extreme outliers. The majority of penguins fall within a typical range, which suggests that the data is consistent and does not contain erroneous entries for this variable.

## Limitations

While this dataset is high quality it has limitations in addressing my original question like

Geographic Scope: It only covers three islands in the Palmer Archipelago. We cannot extrapolate these findings to penguin populations in Antarctica at large or in other regions of the world.

Temporal Scope: The data was collected between 2007 and 2009. It does not allow us to see long-term trends, such as how climate change might be affecting body mass over decades, but that's ok by me

## Follow on

Based on this initial look I had here are a few alternate questions to guide future analysis:

Predictive: Can we build a reliable classification model to predict a penguin's species based solely on its flipper length and bill depth?

Ecological: Is there a correlation between the year of observation and average body mass, potentially hinting at environmental changes during those three years?

Dimorphism: How pronounced is sexual dimorphism (size differences between males and females) across the three different species?

###### AI USAGE

Shotout to the coding assistan within R (copilot I believe) for helping me with the wording of the analysis.