---
title: "Building Footprints by State"
author: "Owen Morales"
format:
  html:
    code-fold: true
    code-line-numbers: true
execute:
  error: true
---

## Libraries

```{r message=FALSE, warning=FALSE}

library(tidyverse)
library(sf)
library(USAboundaries)
library(buildings)
library(geofacet)

```

## Data Wrangling

```{r}

statesSf <- us_states() %>%
  filter(!(state_abbr %in% c("PR", "VI", "GU", "AS", "MP"))) 

idahoCountiesSf <- us_counties(states = "Idaho")

statePermits <- permits %>%
  filter(variable == "Single Family") %>%
  group_by(StateAbbr, year) %>%
  summarize(totalPermits = sum(value, na.rm = TRUE), .groups = "drop")

stateSpatial <- statesSf %>%
  left_join(statePermits, by = c("state_abbr" = "StateAbbr"))

idahoPermits <- permits %>%
  filter(StateAbbr == "ID", variable == "Single Family") %>%
  mutate(
    stateFipsStr = sprintf("%02d", state),
    countyFipsStr = sprintf("%03d", county),
    fipsFull = paste0(stateFipsStr, countyFipsStr)
  )

idahoCountiesSf <- idahoCountiesSf %>%
  mutate(fipsFull = paste0(statefp, countyfp))

countySpatial <- idahoCountiesSf %>%
  left_join(idahoPermits, by = "fipsFull")

```

## Graphs

```{r}

nationalPlot <- ggplot(stateSpatial, aes(x = year, y = totalPermits)) +
  geom_line(color = "#2c3e50", linewidth = 1) +
  annotate("rect", xmin = 2006, xmax = 2010, ymin = -Inf, ymax = Inf, alpha = 0.2, fill = "red") +
  facet_geo(~ state_abbr, grid = "us_state_grid2", scales = "free_y") +
  scale_x_continuous(breaks = c(1980, 1995, 2010), labels = c("'80", "'95", "'10")) +
  theme_minimal() +
  labs(title = "The Geographic Spread of the Housing Collapse", x = NULL, y = "Total Permits") +
  theme(
    axis.text.x = element_text(size = 8, angle = 45, hjust = 1),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.minor = element_blank()
  )

print(nationalPlot)

```

```{r}

idahoPlot <- ggplot(countySpatial, aes(x = year, y = value)) +
  geom_line(color = "#18bc9c", linewidth = 1) +
  geom_vline(xintercept = 2007, linetype = "dashed", color = "red", alpha = 0.6) +
  facet_wrap(~ name, scales = "free_y", ncol = 6) +
  scale_x_continuous(breaks = c(1980, 1995, 2010), labels = c("'80", "", "'10")) +
  theme_minimal() +
  labs(
    title = "Single-Family Building Permits Across Idaho Counties", x = "Year", y = "Permits Issued (Independent Scales)") +
  theme(
    axis.text.x = element_text(size = 9, angle = 45, hjust = 1),
    axis.text.y = element_blank(), 
    strip.text = element_text(size = 10, face = "bold"),
    panel.grid.minor = element_blank()
  )

print(idahoPlot)

```

## Summary

The data shows exactly how bad the 2007–2010 housing collapse really was. It wasn't just big cities that got hit; it happened everywhere. If you look at the US map, almost every state saw a massive drop in single-family building permits starting around 2006. We see the exact same thing right here in Idaho. It didn't matter if it was a fast-growing place like Ada and Canyon counties or smaller, outdoor-focused spots like Valley and Blaine counties building permits just completely fell off a cliff when the crash started. Building permits are a good early warning sign for the economy. These charts show that new housing basically froze everywhere, often way before regular people really felt the effects of the 2008 financial crisis

To show these trends honestly, I designed the graphics to focus on the shape of the crash rather than the raw numbers. The biggest challenge with this kind of data is the huge difference in scale. For example, if we put California and North Dakota on the exact same scale, California's massive numbers would make North Dakota's line look completely flat and hide their local crash. By letting each state and county have its own vertical scale, it's way easier to focus on when the boom and bust happened. Also, laying out the national data to look like a US map makes it super easy to spot regional patterns, so anyone can quickly find and understand what happened in their specific area

#### AI USAGE

Helped me write the summary (: